Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-19 summary.tex
minor edits to intro

Physically, a short thermal timescale is needed to reduce the  
stabilizing effect of buoynancy forces in a stably stratified
disk. Because vertical shear is weak in thin disks, $t_c\Omega\ll 1$
is generally needed for the VSI to operate, otherwise buoyancy would
be strongly stabilizing. Our primary goal was to quantify this
requirement, thereby provide a simple way to assess 
whether or not the VSI is relevant in realistic disks.

@@

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. 
We mostly focused on vertically isothermal disks as 
this is the relevant case for protoplanetary disks passively
irradiated by its central star \citep{chiang97}. It also permits a
largely analytical discussion. Our analysis is then an extension of the
linear models considered in \cite{lubow93} by adding vertical shear
and (rapid) thermal relaxation. We obtained explicit solutions 
describing the VSI with thermal relaxation in the thin disk limit.

@@

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks. We considered the Minimum Mass Solar Nebula 
(MMSN) described in \cite{chiang10} and   
modeled the thermal timescale $t_c$ based on realistic
disk properties and dust opacity. 
%This results in $t_c$ being a
% function of position and the perturbation lengthscale of interest. 
We find the MMSN is unstable to the VSI 
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. It is possible to have the VSI at smaller
radii, but only on radial lengthscales signficantly shorter
than the disk scaleheight, and such perturbations may
be subject to viscous decay in a real disk.

@@

\subsubsection{Improved thermal evolution} 
We assumed non-adiabatic effects can be represented by a thermal
relaxation timescale $t_c$. For  
application to protoplanetary disks we connected $t_c$ to the disk
properties such as surface density, temperature and opacity, making
$t_c$ depend on position and the perturbation 
lengthscales. One might improve the realism further by generalizing
the opacity model adopted here to be density (and hence height)
dependent.  
%here to be density-dependent,  i.e. $\kappa_d =
%% \kappa_d(\rho, T)$. 
However, going beyond thermal relaxation requires a
proper treatment of radiative diffusion. % The linear problem then
% involves a Laplacian operator and additional boundary
% conditions must be supplemented. 
We will consider these improvements in a
follow-up study. %additional heat source

\subsubsection{Additional physics} 
Depending on where the VSI operates in a
disk, additional effects may become important. For example, in the
outer parts of a protoplanetary disk, where we estimate the VSI 
operates, disk self-gravity (SG) may play a role. Vertical
self-gravity may favor the VSI by increasing $|\p_z\Omega|$ through
enhancing the density stratification 
(Eq. \ref{vertical_shear}). %sg in pert? no because kx>>1. but maybe for kx~1

>>>>

Physically, a short thermal timescale is needed to reduce the  
stabilizing effect of buoyancy forces in a stably stratified
disk. Because vertical shear is weak in thin disks, $t_c\Omega\ll 1$
is generally needed for the VSI to operate, otherwise buoyancy would
be strongly stabilizing. Our primary goal was to quantify this
requirement, thereby provide a simple way to assess 
whether or not the VSI is relevant in realistic disks.

@@

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. 
We mostly focused on vertically isothermal disks as 
this is the relevant case for protoplanetary disks (PPDs) passively
irradiated by its central star \citep{chiang97}. It also permits a
largely analytical discussion. Our analysis is then an extension of the
linear models considered in \cite{lubow93} by adding vertical shear
and (rapid) thermal relaxation. We obtained explicit solutions 
describing the VSI with thermal relaxation in the thin disk limit.

@@

We applied our linear models to assess the applicability of the VSI in
PPDs. We considered the Minimum Mass Solar Nebula 
(MMSN) described in \cite{chiang10} and   
modeled the thermal timescale $t_c$ based on realistic
disk properties and dust opacity. 
%This results in $t_c$ being a
% function of position and the perturbation lengthscale of interest. 
We find the MMSN is unstable to the VSI 
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. It is possible to have the VSI at smaller
radii, but only on radial lengthscales signficantly shorter
than the disk scaleheight, and such perturbations may
be subject to viscous decay in a real disk.

@@

\subsubsection{Improved thermal evolution} 
We assumed non-adiabatic effects can be represented by a thermal
relaxation timescale $t_c$. For  
application to PPDs we connected $t_c$ to the disk
properties such as surface density, temperature and opacity, making
$t_c$ depend on position and the perturbation 
lengthscales. One might improve the realism further by generalizing
the opacity model adopted here to be density (and hence height)
dependent.  
%here to be density-dependent,  i.e. $\kappa_d =
%% \kappa_d(\rho, T)$. 
However, going beyond thermal relaxation requires a
proper treatment of radiative diffusion. % The linear problem then
% involves a Laplacian operator and additional boundary
% conditions must be supplemented. 
We will consider these improvements in a
follow-up study. %additional heat source

\subsubsection{Additional physics} 
Depending on where the VSI operates in a
disk, additional effects may become important. For example, in the
outer parts of a PPD, where we estimate the VSI 
operates, disk self-gravity (SG) may play a role. Vertical
self-gravity may favor the VSI by increasing $|\p_z\Omega|$ through
enhancing the density stratification 
(Eq. \ref{vertical_shear}). %sg in pert? no because kx>>1. but maybe for kx~1
